---
title: "data.table learning note"
output: html_notebook
---

```{r}
library(data.table)
```

```{r}
ir = as.data.table(iris)
ir[1:10]
```
Or create a data.table like this:
```{r}
#characters won't be converted to factors by default
DT = data.table(ID = c("b","b","b","a","a","c"), a = 1:6, b = 7:12, c = 13:18)
```


**Data.table operation is mostly done within the square bracket.**
dataTable[i,j,by]
dataTable[subset, columnOperation, groupBy]


### First argument: subset of data
When condition set to columns, it's equivalent to SQL WHERE clause and dplyr's filter() function.
```{r}
ir[Species == 'setosa' & Sepal.Width ==3.5]
```

When condition set to integers, it's interpreted as row index.
```{r}
ir[2:3]
head(ir[order(Sepal.Length, decreasing = TRUE)])
```

### second argument: selection and operation on columns.

if only one column selected, will default to a vector
```{r}
ir[1:2, Species]
```

or you can turn it into a data.table like this:
```{r}
ir[1:3, .(Species)]
#equivalent to ir[, list(Species)]
```




select multiple columns, and name them
```{r}
ir[1:3, .(Width = Sepal.Width, Species)]
#all columns between Sepal.Width and Species
ir[1:3, Sepal.Width:Species]
#all columns not between Sepal.Width and Species
ir[1:3, !(Sepal.Width:Species)]
```

Or you can refer to column indices:

```{r}
ir[1:2, 1:3]
```



elementwise mutations
```{r}
# adding a new column to [ir]
ir[, LargeSepalWidth := Sepal.Width > 3.2]
ir[1:3]
```


or you can do aggregation functions
```{r}
ir[Species == 'setosa', sum(Sepal.Width > 3.2)]
#this is equivalent to:
ir[Species == 'setosa' & Sepal.Width > 3.2, .N]
```

'with' argument
```{r}
#refering to columns by names
ir[1:3, c('Species', 'Sepal.Width'), with = FALSE]

```


### third argument: by
```{r}
# count of samples of each species with sepal with > 3.2
# equivalent to: SELECT COUNT(*) WHERE Sepal.Width > 3.2 GROUP BY Species
ir[Sepal.Width > 3.2, .N, Species]
# group by multiple variables
ir[, .N, .(Species, LargeSepalWidth)]

#recall: LargeSepalWidth is defined as below
#ir[, LargeSepalWidth := Sepal.Width > 3.2]
```
Actually, the above example can be inplemented in one call.
We can actually do groupings within the by argument
```{r}
ir[, .N ,by = .(Species, Sepal.Width > 3.2)]
```


keyby: 
Equivalent to GROUP BY [KEYS] ORDER BY [KEYS]
```{r}
ir[, .N, keyby = .(Species, LargeSepalWidth)]
```


### chaining
follow a data.table [expression] by another set of [expression]
```{r}
ir[, .N, .(Species, LargeSepalWidth)][order(Species, -LargeSepalWidth)]
```




### Now it's the best part.


Subset of Data: .SD()
.SD refers to the subset. It's a data.table


```{r}
ir[,head(.SD,2),by = Species]
```

```{r}
ir[, print(.SD[1:3,Sepal.Width:Petal.Length]), by = Species]
```

You can do all sorts of fancy stuff on this data.table!


As long as j-expression returns a list, each element of the list will be converted to a column in the resulting data.table??


```{r}
lapply(ir[,.(Sepal.Width)], mean)  #a list of length 1
lapply(ir[, Sepal.Length:Petal.Length], mean)   #a list of length 3
foo(ir)
bas(ir)
```


```{r}
#this is the mean of summation of sepal width and petal length, grouped by species. It totally doesn't have to be done this way..
ir[, print(lapply(.SD[,.(Sepal.Width+Petal.Length)], mean)), by = Species]
#this is the mean for each variable between sepal.length and petal.length, grouped by species
ir[, lapply(.SD[, Sepal.Length:Petal.Length], mean), by = Species]
```



```{r}
foo = function(dt){
  return(list(a = dt[1:2, 1], 
              b = dt[3, 1:2]))
}

bar = function(dt){
  return(dt[1, 1:2])
}

bas = function(dt){
  return(list(a = 1, b = 2))
}

xip = function(dt){
  return(list(a = 1, b = dt[1,1]))
}
```


Each subset should return a list (or a vector)

```{r}
#ir[, .(bar(.SD)), by = Species]  # this won't run, because each subset returns a data.table
a = ir[, .(list(bar(.SD))), by = Species]  #this works
a
```


if each subset returns a list with two items: \$a, \$b
this nested structure won't be preserved. It'll be flatten with this order:
sub1\$a, sub1\$b, sub2\$a, sub2\$b, sub3\$a, sub3\$b

```{r}
b = ir[, .(foo(.SD)), by = Species]
b$V1 # a list of 6 items

ir[, bas(.SD), by = Species]
ir[, foo(.SD), by = Species]

```



As long as j returns a list,

```{r}
ir[, .(val = Sepal.Length), by = Species]
```



```{r}
DT
```

```{r}
DT[, list(c(a,b)), by = ID]
```




```{r}
DT[, .(val = c(a,b)), by = ID]
```



```{r}
ir[, list(list(.SD[1:2, 1], .SD[3, 1:2])), by = Species]
```

